Multi-scale trend analysis of water quality using error propagation of generalized additive models

### 2020-12-03 manu_draft.Rmd
additional refs to discussion

Many environmental monitoring programs generate fine-scale but potentially irregular time-series data to assess long-term trends for regulatory, management, or research purposes. The mismatch between the scales of monitoring versus analysis questions or management goals can lead to statistical challenges [@Urquhart98;@Cumming06;@Forbes18] [Marcus will go over refs]. At finer temporal scales, environmental systems may show short-term fluctuations from multiple factors (e.g., weather events, management, or seasonal changes).  Such fluctuations may not be of direct interest for longer-term trends or may not be well-suited to multi-scale smoothing methods.  However, aggregate features of seasonal patterns that integrate or summarize short-term fluctuations may be of interest.  In this paper, we describe methods to estimate across-year trends of within-year features of interest such as a seasonal average, seasonal peak, or seasonal timing of events, while accounting for uncertainties across analysis steps.

Previous methods for water quality trend analysis can be generalized into four basic approaches.  Seasonal Kendall tests or related non-parametric approaches have been used for decades in water quality trend assessments to identify monotonic changes that account for seasonal variation between years [@Hirsch82;@Helsel20].  A literature survey of @Wan17 revealed non-parametric approaches to be the most commonly used methods in long-term water quality trend analysis, yet they have limited scope.  They do not account for changes occurring at different temporal scales, do not adequately evaluate irregularly spaced data [is this true? I would think so.], do not incorporate other explanatory variables, and do not estimate a model that could be useful for other purposes.  Thus, while these non-parametric approaches have some degree of robustness, they apply only to narrow goals.

@@

We illustrate the proposed methods with the motivating example of water quality monitoring in the southern portion of San Francisco Bay, California, USA.  For several decades, approximately twice-monthly monitoring has been conducted at fixed locations (stations) of the longitudinal axis of the Bay.  Analysis of these data is complicated by irregularities in timing and consistency of data collection, such that simple seasonal averages of raw data may not adequately describe trends.  Examples of long-term trend questions include: Are there significant trends in spring mean chlorophyll at four-year (or other) time-scales?  At what across-year scale do within-year summer-fall mean chlorophyll levels change?  Are there significant across-year trends in within-year timing of the spring bloom in chlorophyll or in baseline levels during periods of low productivity?  We also provide an approach for using meta-analysis methods [@Gasparrini12;@Sera19] following signal extraction with GAMs that is new to environmental trend-detection problems.  For this step, we give methods for isolating seasonal trends secondarily from GAM results with reasonable certainty and evaluating these trends between years.

@@

The distinct aspect of model SYDI is the anticipation that within-year fluctuations will vary smoothly across years.  For example, if one year has an early and large magnitude spring bloom, the years before and after would have similar patterns described by the model, perhaps different but not substantially. For the raw San Francisco Bay data, this may not be the case, as one year might have an early, high spring bloom and the next year a late, low spring bloom, or other such large variation across years.  If the interaction spline has sufficient knots, then the degree of smoothness across years can be estimated to be very low, allowing fluctuations within years to vary strongly across years.  However, if that is the case, the conceptual motivation for model SYDI may not be justified.  In summary, model SDYI may best be suited for systems where within-year fluctuations actually vary smoothly across years, in which case it may out-perform the other models.  If that is not the case, it may perform similarly to the others, but the spline structure and penalty structure are more different, making performance harder to compare (Table \@ref(tab:modstrtab)).

@@

We present a simple visual comparison of chlorophyll estimates from models SY, SYD, and SYDI to emphasize that similar fits can be achieved by all of the presented models (Figure \@ref(fig:modsumfig), S is identical to SY and is not shown).  Models SY, SYD, and SYDI were fit to chlorophyll data from station 34 using large k values for the arguments `num_knots_y`, `num_knots_D`, `num_knots_Y_ti`, and `num_knots_D_ti` for each model.  Model predictions by day of year are visually similar between the models (Figure \@ref(fig:modsumfig)a) and are very similar when compared to a 1:1 line (Figure \@ref(fig:modsumfig)b).  However, when comparing similarities among the models using only the continuous year smoother (`s(cont_year)`, as an example), the models differ substantially (Figure \@ref(fig:modsumfig)) because of structural differences and differences in the penalties applied to the basis functions.  These results are also demonstrated in differences between the effective degrees of freedom among the additive components of each model in Table \@ref(tab:modstrtab). Accordingly, the models may differ in which structural component describes variation in the chlorophyll time series but the model predictions are very similar.

@@

This section explains how the within-year features ($\hmt$) from the estimated GAMs, and their standard errors $\hshmt$, are used to investigate across-year trends of within-year features.  In the examples, we use two seasonal averages of chlorophyll as features, one for January-July and the other for August-December.  These periods are chosen for their relevance to phytoplankton bloom phenology and broad seasonal patterns that occur in the San Francisco Bay region. For any feature of interest that includes estimates with associated standard errors, meta-analysis methods can be used to address two kinds of questions:

@@

To test the null hypothesis that  $\mu_t = \mu_r$ for years $r$ and $t$, one can use simple normal distribution theory that $\hmt - \hat{\mu}_r$ has standard error $(\hshmt^2 + \hshmr^2)^{1/2}$.  To test the null hypothesis that all means in a group of (>2) years are equal, one can use a simple case of the meta-analysis given in the next section or a general linear hypothesis test. The latter readily includes consideration of multiple testing, which is provided in the R package `multcomp` [@Hothorn08].  While this package typically uses the results of a linear or generalized linear model fit as input, alternative values of $\hmt$ and $\hshmt$ could be provided as inputs.  Details are given in the Appendix.

@@

The R package *mixmeta* was used to estimate the model in equation \@ref(eqn:mixmet) [@Sera19].  Results have a similar interpretation as those from regression analysis, but parameter estimates and their standard errors incorporate the known standard errors of the response values.  Following meta-analysis theory, $mixmeta$ evaluates this model as a linear mixed effects model with some variance components fixed, whereas others are estimated ($\sbs$).  The default estimation method for meta-analysis models is restricted maximum likelihood (REML).

@@

For all results, model S was used with a sufficiently high number of knots in `num_knots_y` to evaluate chlorophyll trends across the monitoring stations in South San Francisco Bay.  This model was chosen because of the relatively quicker processing time to fit the model, while providing nearly identical explanatory power as compared to the other models (Table \@ref(tab:modstrtab)). Chlorophyll trends were generally well-explained across all stations (Table \@ref(tab:modprftab)), with an average R-squared value equal to `r aver2`% and ranging from `r minr2`% (station `r minst`) to `r maxr2`% (station `r maxst`). GAM predictions from north to south on the longitudinal axis provided different descriptions of annual and seasonal changes in chlorophyll (Figure \@ref(fig:prddat)). All models suggested an increase in chlorophyll in the period of record from 1990 to approximately 2005 to 2010, followed by a decreasing trend until the end of the record. Seasonally, all models indicated a larger spring peak in chlorophyll, particularly for more southern stations, whereas a subsequent peak in fall concentrations did not have any noticeable difference in magnitude by location (Figure \@ref(fig:prddat)).

@@

Our example in Figures \@ref(fig:trndcmpex) and \@ref(fig:trndcmp) demonstrate how different conclusions can be obtained if propagation of uncertainty from raw observations across methods is unaccounted for in trend assessment.  These comparisons are analogous to the example in the previous paragraph where a common approach prior to trend analysis is to aggregate raw observations.  Our assessment of trends using linear regression applied to seasonal averages from the raw observations is effectively similar to averaging results within a year and applying a simple Kendall test. In many cases the results may be similar, but loss of information can lead to increased Type I or II error rates depending on characteristics of the raw data.  This is particularly problematic if sampling effort varies substantially between years.  Averages may be skewed in a particular direction if an annual estimates are based on a handful of observations from select months (e.g., summer only)[@Fouquet12].  The use of GAMs to fit the long-term trend will reduce the potential of limited observations in a particular year skewing estimates of annual or seasonal averages. More importantly, limited observations in a year will be reflected in the standard error estimates derived from the GAM, which has direct implications for how uncertainty is treated in the mixed-meta regression analyses.  As a result, trend assessments from mixed-meta regression should be considered an accurate representation of the actual trend within the limits of the raw observations.

The underlying cross-validation methods used by GAMs in the mgcv package also reduce the decisions required by an analyst that may be necessary for the implementation of alternative trend assessment methods.  For example, WRTDS and similar smoothing approaches (e.g., LOESS) require decisions on appropriate window widths or bandwidths to define the neighborhood of observations for smoothing.  Prior to the implementation of generalized cross-validation in GAMs, these decisions were also required, leading to potential under- or over-smoothing of fitted results to the raw observations.  Because these decisions are no longer needed for GAMs, the results can be considered a more objective and potentially "true" signal of actual trends that are minmally influenced process or observation error present in the raw data.  However, we acknowledge that the third stage analyses that require explicit user inputs on year periods to define trends partially impose arbitrary decisions that can influence interpretation of results.  Although there are undoubtedly many scenarios where years of interest can be chosen objectively (e.g., regulatory compliance periods, time since management intervention, etc.), a more general question of "when" changes occur independent of user decisions is also important to address.  For example, trends could be assessed based on five or ten year moving windows, but which result should be used to inform decisions?  There may be additional methods to develop using objective criteria to more accurately identify inflection points or other important periods where changes occur independent of a user choice.

The trend assessment approach also does not address causality in the observed trends.  In this regard, the approach is similar to other trend assessment methods where the focus is understanding direction, magnitude, and confidence in changes in water quality variables over time. However, a logical follow-up question is "what" is driving the trend after the trend has been adequately described.  This information has obvious implications for management decisions on factors that influence water quality changes, e.g., wastewater treatment upgrades, large-scale climatic factors, or flow regulation practices.  An advantage of GAMs is their flexibility in including alternative predictors, such that the significance of a predictor or comparison of nested models with and without different predictors can provide evidence of which predictors are driving the observed trends. In such cases, considerations of model structure have direct implications on conclusions regarding potential causality.  As noted above, model structure in describing the long-term trend component was irrelevant for describing long-term trends, although a distinction here is made in the objective of the analysis.  Our goal herein was to describe chlorophyll changes relative to time where the predictors were variations on a general theme (e.g., season vs. year), whereas using GAMs with different predictors assess potential causality is a different application with alternative goals.  Therefore, using our approach to assess causality will require considerations of model structure given how GAMs could be used to assess different questions.

Finally, the evaluation of alternative water quality variables in addition to chlorophyll is a trivial but logical extension of the methods proposed in this study.  The long-term monitoring program maintained by USGS includes multiple parameters in addition to chlorophyll that can provide additional context into broader water quality trends in South Bay. These parameters include salinity, temperature, light attenuation, dissolved oxygen, suspended particulate matter, and dissolved inorganic nutrients, which collectively can be used to provide a broader understanding of potential eutrophication patterns or ecosystem shifts at seasonal and multi-decadal scales.  Measurements of dissolved oxygen can also be used to estimates gross primary production to assess process rates that may be more indicative of system function.  We provide an open-source R package to facilitate these analyses (see supplement) that includes all methods described in this manuscript.

@@

When `select = TRUE` is included, the comparison between models S and SY changes.  This option tells `mgcv` to penalize the coefficient of the linear terms in the spline.  This would be appropriate if `cont_year` was an explanatory variable subject to variable selection, but it would be a strange choice if one is explicitly including a linear term for `cont_year`.  If `select = TRUE` is used, models S and SY would still be effectively equivalent, but it would appear by AIC selection that one model is superior.  That would be an artifact of the choice in model SY to include a linear trend in `cont_year` both as a separate term and as part of the spline, with the latter subject to penalization.  A link to an RMarkdown file demonstrating this issue is provided.

@@

# table stuff
cap.val <- 'Comparison of the four model structures (S, SY, SYD, SYDI) described in the first stage analysis of GAM estimation.  The four models provide either identical or comparable ability to describe chlorophyll trends at an example station (32) in South San Francisco Bay.  The models differ in additive smoothers and the amount of effective degrees of freedom (edf) in the smoothers (measure of wiggliness in each component), but the overall model predictions are comparable. GCV: generalized cross-validation score, R2: r-squared values for predictions, edf: effective degrees of freedom, F: F-statistic, p-val: probability value, ** p < 0.001'

>>>>

Many environmental monitoring programs generate fine-scale but potentially irregular time-series data to assess long-term trends for regulatory, management, or research purposes. The mismatch between the scales of monitoring versus analysis questions or management goals can lead to statistical challenges [@Urquhart98;@Cumming06;@Forbes18]. At finer temporal scales, environmental systems may show short-term fluctuations from multiple factors (e.g., weather events, management, or seasonal changes).  Such fluctuations may not be of direct interest for longer-term trends or may not be well-suited to multi-scale smoothing methods.  However, aggregate features of seasonal patterns that integrate or summarize short-term fluctuations may be of interest.  In this paper, we describe methods to estimate across-year trends of within-year features of interest such as a seasonal average, seasonal peak, or seasonal timing of events, while accounting for uncertainties across analysis steps.

Previous methods for water quality trend analysis can be generalized into four basic approaches.  Seasonal Kendall tests or related non-parametric approaches have been used for decades in water quality trend assessments to identify monotonic changes that account for seasonal variation between years [@Hirsch82;@Helsel20].  A literature survey of @Wan17 revealed non-parametric approaches to be the most commonly used methods in long-term water quality trend analysis, yet they have limited scope.  For descriptive decomposition of long-term monitoring data, they do not account for changes occurring at different temporal scales, do not adequately evaluate irregularly spaced data, do not incorporate other explanatory variables, and do not estimate a model that could be useful for other purposes.  Thus, while these non-parametric approaches have some degree of robustness to assess magnitude and direction of trends, they apply only to narrow goals.

@@

We illustrate the proposed methods with the motivating example of water quality monitoring in the southern portion of San Francisco Estuary, California, USA.  For several decades, approximately twice-monthly monitoring has been conducted at fixed locations (stations) of the longitudinal axis of the Bay.  Analysis of these data is complicated by irregularities in timing and consistency of data collection, such that simple seasonal averages of raw data may not adequately describe trends.  Examples of long-term trend questions include: Are there significant trends in spring mean chlorophyll at four-year (or other) time-scales?  At what across-year scale do within-year summer-fall mean chlorophyll levels change?  Are there significant across-year trends in within-year timing of the spring bloom in chlorophyll or in baseline levels during periods of low productivity?  We also provide an approach for using meta-analysis methods [@Gasparrini12;@Sera19] following signal extraction with GAMs that is new to environmental trend-detection problems.  For this step, we give methods for isolating seasonal trends secondarily from GAM results with reasonable certainty and evaluating these trends between years.

@@

The distinct aspect of model SYDI is the anticipation that within-year fluctuations will vary smoothly across years.  For example, if one year has an early and large magnitude spring bloom, the years before and after would have similar patterns described by the model, perhaps different but not substantially. For the raw SFE data, this may not be the case, as one year might have an early, high spring bloom and the next year a late, low spring bloom, or other such large variation across years.  If the interaction spline has sufficient knots, then the degree of smoothness across years can be estimated to be very low, allowing fluctuations within years to vary strongly across years.  However, if that is the case, the conceptual motivation for model SYDI may not be justified.  In summary, model SDYI may best be suited for systems where within-year fluctuations actually vary smoothly across years, in which case it may out-perform the other models.  If that is not the case, it may perform similarly to the others, but the spline structure and penalty structure are more different, making performance harder to compare (Table \@ref(tab:modstrtab)).

@@

We present a simple visual comparison of chlorophyll estimates from models SY, SYD, and SYDI to emphasize that similar fits can be achieved by all of the presented models (Figure \@ref(fig:modsumfig), S is identical to SY and is not shown).  Models SY, SYD, and SYDI were fit to chlorophyll data from station 34 using large k values for the arguments `num_knots_y`, `num_knots_D`, `num_knots_Y_ti`, and `num_knots_D_ti` for each model.  Model predictions by day of year are visually similar between the models (Figure \@ref(fig:modsumfig)a) and are very similar when compared to a 1:1 line (Figure \@ref(fig:modsumfig)b).  However, when comparing similarities among the models using only the continuous year smoother (`s(cont_year)`, as an example), the models differ substantially because of structural differences and differences in the penalties applied to the basis functions.  These results are also demonstrated in differences between the effective degrees of freedom among the additive components of each model in Table \@ref(tab:modstrtab). Accordingly, the models may differ in which structural component describes variation in the chlorophyll time series, but the model predictions are very similar.

@@

This section explains how the within-year features ($\hmt$) from the estimated GAMs, and their standard errors $\hshmt$, are used to investigate across-year trends of within-year features.  In the examples, we use two seasonal averages of chlorophyll as features, one for January-July and the other for August-December.  These periods are chosen for their relevance to phytoplankton bloom phenology and broad seasonal patterns that occur in the SFE region. For any feature of interest that includes estimates with associated standard errors, meta-analysis methods can be used to address two kinds of questions:

@@

To test the null hypothesis that  $\mu_t = \mu_r$ for years $r$ and $t$, one can use simple normal distribution theory that $\hmt - \hat{\mu}_r$ has standard error $(\hshmt^2 + \hshmr^2)^{1/2}$.  To test the null hypothesis that all means in a group of (>2) years are equal, one can use a simple case of the meta-analysis given in the next section or a general linear hypothesis test. The latter readily includes consideration of multiple testing, which is provided in the R package `multcomp` [@Hothorn08].  While this package typically uses the results of a linear or generalized linear model fit as input, alternative values of $\hmt$ and $\hshmt$ could be provided as inputs.

@@

The R package *mixmeta* was used to estimate the model in equation \@ref(eq:mixmet) [@Sera19].  Results have a similar interpretation as those from regression analysis, but parameter estimates and their standard errors incorporate the known standard errors of the response values.  Following meta-analysis theory, $mixmeta$ evaluates this model as a linear mixed effects model with some variance components fixed, whereas others are estimated ($\sbs$).  The default estimation method for meta-analysis models is restricted maximum likelihood (REML).

@@

For all results, model S was used with a sufficiently high number of knots in `num_knots_y` to evaluate chlorophyll trends across the monitoring stations in South SFE.  This model was chosen because of the relatively quicker processing time to fit the model, while providing nearly identical explanatory power as compared to the other models (Table \@ref(tab:modstrtab)). Chlorophyll trends were generally well-explained across all stations (Table \@ref(tab:modprftab)), with an average R-squared value equal to `r aver2`% and ranging from `r minr2`% (station `r minst`) to `r maxr2`% (station `r maxst`). GAM predictions from north to south on the longitudinal axis provided different descriptions of annual and seasonal changes in chlorophyll (Figure \@ref(fig:prddat)). All models suggested an increase in chlorophyll in the period of record from 1990 to approximately 2005 to 2010, followed by a decreasing trend until the end of the record. Seasonally, all models indicated a larger spring peak in chlorophyll, particularly for more southern stations, whereas a subsequent peak in fall concentrations did not have any noticeable difference in magnitude by location (Figure \@ref(fig:prddat)).

@@

Our example in Figures \@ref(fig:trndcmpex) and \@ref(fig:trndcmp) demonstrate how different conclusions can be obtained if propagation of uncertainty from raw observations across methods is unaccounted for in trend assessment.  These comparisons are analogous to the example in the previous paragraph where a common approach prior to trend analysis is to aggregate raw observations.  Our assessment of trends using linear regression applied to seasonal averages from the raw observations is effectively similar to averaging results within a year and applying a simple Kendall test. In many cases the results may be similar, but loss of information can lead to increased Type I or II error rates depending on characteristics of the raw data.  This is particularly problematic if sampling effort varies substantially between years.  Averages may be skewed in a particular direction if an annual estimates are based on a handful of observations from select months (e.g., summer only)[@Fouquet12].  The use of GAMs to fit the long-term trend will reduce the potential of limited observations in a particular year skewing estimates of annual or seasonal averages. More importantly, limited observations in a year will be reflected in the standard error estimates derived from the GAM, which has direct implications for how uncertainty is treated in the mixed-meta regression analyses [@Sera19].  As a result, trend assessments from mixed-meta regression should be considered an accurate representation of the actual trend within the limits of the raw observations.

The underlying cross-validation methods used by GAMs in the mgcv package also reduce the decisions required by an analyst that may be necessary for the implementation of alternative trend assessment methods.  For example, WRTDS and similar smoothing approaches (e.g., LOESS) require decisions on appropriate window widths or bandwidths to define the neighborhood of observations for smoothing [@Hirsch10;@Wan17].  Prior to the implementation of generalized cross-validation in GAMs, these decisions were also required, leading to potential under- or over-smoothing of fitted results to the raw observations.  Because these decisions are no longer needed for GAMs, the results can be considered a more objective and potentially "true" signal of actual trends that are minimally influenced process or observation error present in the raw data.  However, we acknowledge that the third stage analyses that require explicit user inputs on year periods to define trends partially impose arbitrary decisions that can influence interpretation of results.  Although there are undoubtedly many scenarios where years of interest can be chosen objectively (e.g., regulatory compliance periods, time since management intervention, etc.), a more general question of "when" changes occur independent of user decisions is also important to address.  For example, trends could be assessed based on five or ten year moving windows, but which result should be used to inform decisions?  There may be additional methods to develop using objective criteria to more accurately identify inflection points or other important periods where changes occur independent of a user choice.

The trend assessment approach also does not address causality in the observed trends.  In this regard, the approach is similar to other trend assessment methods where the focus is understanding direction, magnitude, and confidence in changes in water quality variables over time. However, a logical follow-up question is "what" is driving the trend after the trend has been adequately described.  This information has obvious implications for management decisions on factors that influence water quality changes, e.g., wastewater treatment upgrades, large-scale climatic factors, or flow regulation practices.  An advantage of GAMs is their flexibility in including alternative predictors, such that the significance of a predictor or comparison of nested models with and without different predictors can provide evidence of which predictors are driving the observed trends [@Wood02;@Zuur09]. In such cases, considerations of model structure can have direct implications on conclusions regarding potential causality.  As noted above, model structure in describing the long-term trend component was irrelevant for describing long-term trends, although a distinction here is made in the objective of the analysis.  Our goal herein was to describe chlorophyll changes relative to time where the predictors were variations on a general theme (e.g., season vs. year), whereas using GAMs with different predictors assess potential causality is a different application with alternative goals.  Therefore, using our approach to assess causality will require considerations of model structure given how GAMs could be used to assess different questions.

Finally, the evaluation of trends for alternative water quality variables in addition to chlorophyll is a trivial but logical extension of the methods proposed in this study.  The long-term monitoring program maintained by USGS includes multiple parameters in addition to chlorophyll that can provide additional context into broader water quality trends in South Bay [@Cloern16;@Schraga20]. These parameters include salinity, temperature, light attenuation, dissolved oxygen, suspended particulate matter, and dissolved inorganic nutrients, which collectively can be used to provide a broader understanding of potential eutrophication patterns or ecosystem shifts at seasonal and multi-decadal scales.  Chlorophyll measurements can also be used to estimates gross primary production to assess process rates that may be more indicative of system function [@Jassby02;@Cloern07].  We provide an open-source R package to facilitate these analyses (see supplement) that includes all methods described in this manuscript.

@@

A note about the `select` argument in `mgcv`: When `select = TRUE` is included, the comparison between models S and SY changes.  This option tells `mgcv` to penalize the coefficient of the linear terms in the spline.  This would be appropriate if `cont_year` was an explanatory variable subject to variable selection, but it is irrelevant if including both a linear and spline term for `cont_year`.  If `select = TRUE` is used, models S and SY would still be effectively equivalent, but AIC selection would suggest that one model is superior.  This result would be an artifact of the choice in model SY to include a linear trend in `cont_year` both as a separate term and as part of the spline, with the latter subject to penalization.

@@

# table stuff
cap.val <- 'Comparison of the four model structures (S, SY, SYD, SYDI) described in the first stage analysis of GAM estimation.  The four models provide either identical or comparable ability to describe chlorophyll trends at an example station (32) in South San Francisco Estuary.  The models differ in additive smoothers and the amount of effective degrees of freedom (edf) in the smoothers (measure of wiggliness in each component), but the overall model predictions are comparable. GCV: generalized cross-validation score, R2: r-squared values for predictions, edf: effective degrees of freedom, F: F-statistic, p-val: probability value, ** p < 0.001'
